\documentclass[10.5pt, a4paper, oneside]{report}
\usepackage{layout}
\usepackage[margin=102pt, textwidth=390pt]{geometry}

\usepackage{polyglossia}
\setmainlanguage{english}
\setotherlanguage{german}
\usepackage{fontspec}
\setmainfont{TeX Gyre Pagella}
\setsansfont{Fira Sans}[Scale=MatchLowercase]
\setmonofont{Inconsolata}[Scale=MatchLowercase]
\usepackage{sectsty}
\allsectionsfont{\sffamily}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
\DeclarePairedDelimiter\nint{\lfloor}{\rceil}
\def\RR{\mathbb{R}}
\def\Pr{\mathrm{Pr}}
\def\NN{\mathbb{N}}
\def\im{\mathrm{im}}
\def\cond{\,|\,}
\def\sgn{\mathrm{sgn}}
\def\lamishigh{$\text{LAMIS}_{\text{high}}$}
\def\lamislow{$\text{LAMIS}_{\text{low}}$}
\def\tbeta{\tilde{\beta}}
\def\hbeta{\hat{\beta}}

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup[table]{position=bottom}
\captionsetup[subtable]{font=normalsize}

\usepackage{siunitx}

\usepackage{natbib}
\bibliographystyle{abbrvnat}
\setcitestyle{numbers,square}

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}

\begin{document}
    \begin{titlepage}
        \centering
        \vspace*{1cm}
        
        \huge
        \textsf{
        \textbf{Computational detection of a high-risk DLBCL group}
        }
        
        \vspace{1.5cm}
        
        \large
        \textbf{Lukas Geßl}
        
        \vfill
        
        \large
        A thesis presented for the degree of\\
        Master of Science
        
        \vspace{0.8cm}
        
        \begin{tabular}{rl}
            First Supervisor: & Prof.\ Dr.\ Harald Garcke \\
            Second Supervisor: & Prof.\ Dr.\ Rainer Spang \\
        \end{tabular}
        
        \vspace{0.8cm}
        
        Department of Mathematics\\
        University of Regensburg\\
        \vspace{0.8cm}
        \today
        
    \end{titlepage}

    \begin{abstract}
        Chemotherapy with R-CHOP is the standard treatment for diffuse large B-cell lymphoma, 
        the most common type of non-Hodgkin lymphoma, curing about two thirds of 
        patients. Survival for the remaining third with refractory or relapsed disease, however, 
        remains poor. Pharma-sponsored randomized trials in the whole DLBCL population to date have 
        failed 
        to improve R-CHOP. The International Prognostic Index (IPI), the only widely accepted 
        risk-assessment tool for DLBCL and an easy clinical test, fails to identify 
        a high-risk DLBCL subpopulation that is 
        large and precise enough to trigger research and enable clinical trials for new treatments 
        that outperform R-CHOP on this subpopulation. 

        This thesis aims to develop a computational method that identifies DLBCL patients with 
        progression-free survival (PFS) below two years with higher prevalence and significantly 
        higher precision than the IPI and to show this on independent data. It also deals with the 
        question under which circumstances we can do so reliably.

        After introducing into DLBCL, the IPI together with its shortcomings and MMML-Predict, the 
        project that this thesis is part of, in the first chapter, we describe the statistical 
        frame and models as well as our software in the second chapter. The third chapter applies 
        the method to three different data sets and one big data set comprised of these three 
        data sets and 
        shows that we can indeed deliver a model with the desired properties. Analysis after 
        freezing the models and unlocking the test data suggest that, for a reliable internal 
        validation and high test performance, the following factors are crucial:
        (a) data sets with a large number of samples, even if 
        they result from combining somewhat different, partly non-prospective data sets, 
        (b) relying on already-existing molecular signatures rather than fitting new ones and 
        (c) deploying simple, generalized linear models that can handle batch effects. 
        The fourth chapter 
        discusses how we can develop even better models in the future, especially for MMML-Predict. 

        We conclude that with currently available data and statistical and computational methods, 
        we can identify a DLBCL subpopulation with poor survival that is larger and 
        holds significantly more high-risk patients than that identified by the IPI. With more 
        data and new and 
        more accurately measured features in the future, we expect to be able to further improve the 
        performance of such models.
    \end{abstract}

    % \layout
    \tableofcontents
    \input{1-intro.tex}
    \input{2-methods.tex}
    \input{3-results.tex}
    \input{4-discussion.tex}
    \bibliography{../lit.bib}

    \newpage
    \section*{Declaration of authorship}
    \begin{german}
        Ich habe die Arbeit selbständig verfasst, keine anderen als die angegebenen Quellen und 
        Hilfsmittel benutzt und bisher keiner anderen Prüfungsbehörde vorgelegt. Außerdem 
        bestätige ich hiermit, dass die vorgelegten Druckexemplare und die vorgelegte elektronische 
        Version der Arbeit identisch sind, dass ich über wissenschaftlich korrektes Arbeiten und 
        Zitieren aufgeklärt wurde und dass ich von den in § 26 Abs.\ 5 vorgesehenen Rechtsfolgen 
        Kenntnis habe. 
    \end{german}

        \vspace{1.5cm}
        \noindent Regensburg, \today \hfill \underline{\hspace{6cm}}

\end{document}